Type theory in type theory with single substitutions

### 2024-09-14 types_2024_postproc/p.tex
citations

Kaposi and Luksa \cite{} defined a telescopic single substitution
calculus for simple type theory, it can be seen as a simply typed
version of B-systems. They show that the category of contextual models
of their calculus is equivalent to the category of contextual simply
typed CwFs with function space. Their equivalence result holds even
without the presence of function space. In the presence of function
space, the same equivalence holds for the simply typed version of our
substitution calculus, see the formalisation accompanying this paper.

Most formalisations of type theory are extrinsic
\cite{pujet,abel,metacoq}: they define the syntax as abstract syntax
trees, and equip it with typing and conversion relations. A higher
level representation is working in setoid hell
\cite{chapman09eatitself}: terms are intrinsically well-typed, but
conversion is still an explicit relation. For formalising the
GAT-level syntax, one needs a stronger metatheory than ordinary Agda
or Coq: the metatheory has to support quotient inductive-inductive
types (QIITs \cite{DBLP:journals/pacmpl/KaposiKA19}), in other words,
initial models of GATs. Altenkirch and Kaposi
\cite{DBLP:conf/popl/AltenkirchK16} formalised the syntax of type
theory using postulated QIITs in Agda, together with parametricity and
normalisation \cite{lmcs:4005} proofs. Brunerie and de Boer
\cite{initiality-agda} constructed the initial contextual category in
Agda using postulated quotients. QIITs are supported by the cubical
set model \cite{} and the setoid model \cite{kaposi-qiit-setoid}, thus
Cubical Agda \cite{} and setoid/observational type theory
\cite{setoid,loicThesis} support QIITs. For example, in Cubical Agda,
set-truncated and groupoid-truncated syntaxes of type theory have been
formalised \cite{cohtt}. Without QIITs, tricks such as shallow
embedding the syntax can be used to formalise metatheoretic results
about type theory such as gluing
\cite{kaposi_et_al:LIPIcs:2019:10532}, and its special cases
canonicity and parametricity \cite{kaposi-shallow}.

>>>>

Kaposi and Luksa \cite{luksa} defined a telescopic single substitution
calculus for simple type theory, it can be seen as a simply typed
version of B-systems. They show that the category of contextual models
of their calculus is equivalent to the category of contextual simply
typed CwFs with function space. Their equivalence result holds even
without the presence of function space. In the presence of function
space, the same equivalence holds for the simply typed version of our
substitution calculus, see the formalisation accompanying this paper.

Most formalisations of type theory are extrinsic
\cite{DBLP:journals/pacmpl/0001OV18,DBLP:conf/cpp/AdjedjLMPP24,DBLP:journals/jar/SozeauABCFKMTW20}:
they define the syntax as abstract syntax trees, and equip it with
typing and conversion relations. A higher level representation is
working in setoid hell \cite{chapman09eatitself}: terms are
intrinsically well-typed, but conversion is still an explicit
relation. For formalising the GAT-level syntax, one needs a stronger
metatheory than ordinary Agda or Coq: the metatheory has to support
quotient inductive-inductive types (QIITs
\cite{DBLP:journals/pacmpl/KaposiKA19}), in other words, initial
models of GATs. Altenkirch and Kaposi
\cite{DBLP:conf/popl/AltenkirchK16} formalised the syntax of type
theory using postulated QIITs in Agda, together with parametricity and
normalisation \cite{lmcs:4005} proofs. Brunerie and de Boer
\cite{initiality-agda} constructed the initial contextual category in
Agda using postulated quotients. QIITs are supported by the cubical
set model \cite{DBLP:conf/lics/CoquandHM18} and the setoid model \cite{kaposi-qiit-setoid}, thus
Cubical Agda \cite{DBLP:journals/jfp/VezzosiMA21} and
setoid/observational type theory \cite{setoid,DBLP:phd/hal/Pujet22} support
QIITs. For example, in Cubical Agda, set-truncated and
groupoid-truncated syntaxes of type theory have been formalised
\cite{cohtt}. Without QIITs, tricks such as shallow embedding the
syntax can be used to formalise metatheoretic results about type
theory such as gluing \cite{kaposi_et_al:LIPIcs:2019:10532}, and its
special cases canonicity and parametricity \cite{kaposi-shallow}.
